---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# Welcome to maxbootR

<!-- badges: start -->

<!-- badges: end -->

The `R`-package **maxbootR** provides bootstrap methods for Block Maxima estimators, highly optimized via fast C++ backends. It offers the implementation of the first consistent block maxima bootstrap methods from the paper [Bücher, A. and Staud, T. (2024+)](https://arxiv.org/abs/2409.05529).

## Installation

You can install the development version of maxbootR from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("torbenstaud/maxbootR")
```

or from [CRAN](https://cran.r-project.org/web/packages/maxbootR/index.html) in `R`.

## Example

This is a basic example which shows you how to extract sliding block maxima from data.

```{r blockmax}
library(maxbootR)
# Example: disjoint block maxima
set.seed(91)
blockmax(xx = rnorm(100*365), block_size = 365, type = "sb")
```

The following code bootstraps the 100-year return level of synthetic data worth 90 years and based on a normal distribution for both disjoint and sliding bootstraps. The ratio of variances of both estimation methods is calculated and we see that sliding performs better than disjoint.

```{r bst}
library(maxbootR)
# Example: disjoint block maxima
set.seed(91)
xx <- rnorm(90*365)
bst.db <- 
  maxbootr(
    xx = xx, est = "rl", block_size = 365, 
    B = 500, type = "db", annuity = 100) 
bst.sb <- 
  maxbootr(
    xx = xx, est = "rl", block_size = 365, 
    B = 500, type = "sb", annuity = 100) 

var(bst.sb)/var(bst.db)
```
Test
```{r blockmax-plot, echo = FALSE, fig.align = "center", fig.width = 6, fig.height = 4}
library(ggplot2)
library(maxbootR)
library(dplyr)
# Combine for plotting
df_q <- tibble(
  value = c(bst.db, bst.sb),
  method = c(rep("Disjoint Blocks", length(bst.db)),
             rep("Sliding Blocks", length(bst.sb)))
)

# Histogram plot
ggplot(df_q, aes(x = value)) +
  geom_histogram(fill = "steelblue", color = "white", bins = 30) +
  facet_wrap(~ method, nrow = 1) +
  labs(
    title = "Bootstrap Estimates of 100-Year Return Levels",
    x = "Estimated Return Level",
    y = "Count"
  )
```

# Learn More

Check out the full vignette from the published package: [CRAN VIGNETTE](???)

You'll still need to render `README.Rmd` regularly, to keep `README.md` up-to-date. `devtools::build_readme()` is handy for this.

